Future Evolution of CSMA Protocols for the IEEE 802.11 Standard

### 2013-02-28 paper.tex
working on camera ready version

\documentclass[conference]{IEEEtran}
\usepackage{cite}
% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
\usepackage[linesnumbered,vlined]{algorithm2e}
\usepackage[ansinew]{inputenc}
\usepackage{epsfig}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{subfigure}
\usepackage[usenames]{color}
\usepackage{soul}
\usepackage{comment}
\usepackage{multirow}
\usepackage[cmex10]{amsmath}
\usepackage{bm}
\usepackage{setspace}

@@

\begin{abstract}
%\boldmath
In this paper a candidate protocol to replace the prevalent CSMA/CA medium access control in Wireless Local Area Networks is presented. The proposed protocol can achieve higher throughput compared to CSMA/CA, while maintaining fairness, and without additional implementation complexity. Under certain circumstances, it is able to reach and maintain collision-free operation, even when the number of contenders is variable and potentially large. It is backward compatible, allowing for new and legacy stations to coexist without degrading one another's performance, a property that can make the adoption process by future versions of the standard smooth and inexpensive.

@@

@@

{Note that the coexistence and fairness of CSMA/ECA and CSMA/CA has been studied in \cite{barcelo2010fcc}. The addition of hysteresis and fair-share may represent less frequent transmissions and more packets per transmission, which do not interfere in any way with the execution of the legacy protocol.}

@@

Algorithm \ref{alg:CSMA_ECA} describes the basic CSMA/ECA in which a deterministic backoff is used after successful transmissions. The only change with respect to CSMA/CA is in fact in line $14$, where the random assignment of $b$ in CSMA/CA is replaced by a deterministic assignment in CSMA/ECA. Note that at this point the value of $s$ is zero, and therefore the assigned deterministic value is in fact $b = \text{CW}_{\min}/2$.

Hysteresis is included in Algorithm \ref{alg:CSMA_ECA_hist}. Adding hysteresis is as simple as removing line $16$ from Algorithm \ref{alg:CSMA_ECA}. This means that the deterministic backoff value chosen after a successful transmission in this case is half the length of the backoff stage in which the successful transmission has occurred. In fact, the backoff stage is reset only when the node has no packet to serve. This allows for the use of larger deterministic backoff values, which means that the network can reach a collision-free operation for a higher number of nodes. Since in the collision-free operation, some nodes may have higher backoff values than others, this protocol does not share the medium in a fair manner among nodes.

@@

To evaluate the four protocols, a network of $N$ nodes is considered, where each node is within the coverage area of the others. The nodes are set to be under saturation (always have packets to transmit). The number of contenders, $N$, ranges from $2$ to $50$, and one thousand instances of the simulation are performed for each $N$. All plotted results show the 95\% confidence intervals.

@@

% use section* for acknowledgement

>>>>

\documentclass[conference]{IEEEtran}
\usepackage{cite}
% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
  % \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../pdf/}{../jpeg/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else
  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
  % will default to the driver specified in the system graphics.cfg if no
  % driver is specified.
  % \usepackage[dvips]{graphicx}
  % declare the path(s) where your graphic files are
  % \graphicspath{{../eps/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
  % \DeclareGraphicsExtensions{.eps}
\fi
\usepackage[linesnumbered,vlined]{algorithm2e}
\usepackage{epsfig}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{subfigure}
\usepackage[usenames]{color}
\usepackage{soul}
\usepackage{comment}
\usepackage{multirow}
\usepackage[cmex10]{amsmath}
\usepackage{bm}
\usepackage{setspace}

@@

\begin{abstract}
%\boldmath
In this paper a candidate protocol to replace the prevalent CSMA/CA medium access control in Wireless Local Area Networks is presented. The proposed protocol can achieve higher throughput than CSMA/CA, while maintaining fairness, and without additional implementation complexity. Under certain circumstances, it is able to reach and maintain collision-free operation, even when the number of contenders is variable and potentially large. It is backward compatible, allowing for new and legacy stations to coexist without degrading one another's performance, a property that can make the adoption process by future versions of the standard smooth and inexpensive.

@@

Quality-of-service and traffic differentiation in CSMA/ECA has is discussed in \cite{barcelo2009tpc, he2009srb}.
The performance of CSMA/ECA in the presence of channel errors is considered in \cite{he2009srb, barcelo2010fcc, fang2011dlm, barcelo2012mdc}. 
Other aspects such as coexistence with legacy stations and performance in the presence of new entrants have also been covered in \cite{he2009srb, barcelo2010fcc, fang2011dlm}.
Results for the impact of slot drift (non-ideal clocks) in protocols similar to CSMA/ECA can be found in \cite{gong2012asd}.
In the present workshop paper, we focus on a simple greenfield scenario with a single traffic class, ideal channel, ideal clocks, and saturation conditions.
We are currently evaluating more complex scenarios that we outline in Section \ref{sec:}

@@

Note that the coexistence and fairness of basic CSMA/ECA and CSMA/CA has been studied in \cite{barcelo2010fcc}. 
The addition of hysteresis and fair-share to the basic CSMA/ECA may represent less frequent transmissions and more packets per transmission
These two modifications do not interfere in any way with the execution of the legacy protocol and, therefore, CSMA/CA with hysteresis and fair-share can also coexist with the currently deployed protocol.

@@

Algorithm \ref{alg:CSMA_ECA} describes the basic CSMA/ECA in which a deterministic backoff is used after successful transmissions. The only change with respect to CSMA/CA is in fact in line $18$, where the random assignment of $b$ in CSMA/CA is replaced by a deterministic assignment in CSMA/ECA. Note that at this point the value of $s$ is zero, and therefore the assigned deterministic value is in fact $b = \text{CW}_{\min}/2$.
This value is roughly equal to the expectation of the random backoff used in the same situation in CSMA/CA.
This particular choice guarantees fairness between new stations and legacy stations.

Hysteresis is included in Algorithm \ref{alg:CSMA_ECA_hist}. Adding hysteresis is as simple as removing line $16$ from Algorithm \ref{alg:CSMA_ECA}. This means that the deterministic backoff value chosen after a successful transmission in this case is half the length of the contention window of the backoff stage in which the successful transmission has occurred. In fact, the backoff stage is reset only when the node has no packet to serve. This allows for the use of larger deterministic backoff values, which means that the network can reach a collision-free operation for a higher number of nodes. Since in the collision-free operation, some nodes may have higher backoff values than others, this protocol does not share the medium in a fair manner among nodes.

@@

To evaluate the four protocols, a network of $N$ nodes is considered, where each node is within the coverage area of the others. The channel does not introduce errors and the nodes are set to be under saturation (always have packets to transmit). The number of contenders, $N$, ranges from $2$ to $50$, and one thousand instances of the simulation are performed for each $N$. All plotted results show the 95\% confidence intervals.

@@

\section*{Acknowledgment}

The authors would like to thank the anonymous reviewers for their corrections, helpful comments and suggestions for future work. The authors are also thankful to Alessandro Checco for helping in the preparation of the related work section.
